An Extendable \proglang{Python} Implementation of Robust Optimisation Monte Carlo

### 2021-07-17 latex_files/ch3/3_2_3_inference.tex
updated most content in chapter 2,3

The inference part contains the 4 following functionalities:

\begin{enumerate}
\item \mintinline{python}{romc.sample(n2, seed=None)}
\item \mintinline{python}{romc.compute_expectation(h)}  
\item \mintinline{python}{romc.eval_unnorm_posterior(theta)}
\item \mintinline{python}{romc.eval_posterior(theta)}
\end{enumerate}

\subsubsection*{Function (i): Perform weighted sampling}

\mintinline{python}{romc.sample(n2)}
\vspace{5mm}

\noindent
This is the basic inference utility of the ROMC implementation; we
draw $n_2$ samples for each bounding box region. This gives a total of
$k \times n_2$, where $k < n_1$ is the number of the optimal points
remained after filtering\footnote{From the $n_1$ optimisation
  problems, only the ones with $g_i(\thetab_*) < \epsilon$ are kept
  for building a bounding box}. The samples are drawn from a uniform
distribution $q_i$ defined over the corresponding bounding box and the
weight $w_i$ is computed as in equation~\eqref{eq:sampling}. The
function stores an \pinline{elfi.Result} object as
\pinline{romc.result} attribute. The \pinline{elfi.Result} provides
some usefull functionalities for inspecting the obtained samples e.g.\
\pinline{romc.result.summary()} prints the number of the obtained
samples and their mean. A complete overview of these functionalities is
provided in ELFI's
\href{https://elfi.readthedocs.io/en/latest/api.html#elfi.methods.results.Sample}{official
  documentation}.

\subsubsection*{Function (ii): Compute an expectation}

\mintinline{python}{romc.compute_expectation(h)}
\vspace{5mm}

\noindent
This function computes the expectation
$E_{p(\thetab|\data)}[h(\thetab)]$ using
expression~\eqref{eq:expectation}. The argument \pinline{h} can be
any python \pinline{Callable}.

\subsubsection*{Function (iii): Evaluate the unnormalised posterior}
\mintinline{python}{romc.eval_unorm_posterior(theta, eps_cutoff=False)}
\vspace{5mm}

@@

\subsubsection*{Function (iv): Evaluate the normalised posterior}
\mintinline{python}{romc.eval_posterior(theta, eps_cutoff=False)}
\vspace{5mm}

\noindent
This function evaluates the normalised posterior. For doing so it
needs to approximate the partition function
$Z = \int p_{d,\epsilon}(\thetab|\data)d\thetab$; this is done using
the Riemann integral approximation. Unfortunately, the Riemann
approximation does not scale well in high-dimensional spaces, hence
the approximation is tractable only at low-dimensional parametric
spaces. Given that this functionality is particularly useful for
plotting the posterior, we could say that it is meaningful to be used
for up to $3D$ parametric spaces, even though it is not restricted to
that. Finally, for this functionality to work, the \pinline{bounds}
arguments must have been set at the initialisation of the
\pinline{elfi.ROMC} object.\footnote{The argument \pinline{bounds}
  should define a bounding box \emph{containing} all the mass of the
  prior; it may also contain redundant areas. For example, if the
  prior is the uniform defined over a unit circle i.e.\
  $c=(0,0), r=1$, the best bounds arguments is
  \pinline{bounds=[(-1,1),(-1,1)]}. However, any argument
  \pinline{bounds=[(-t,t),(-t,t)]} where $t\geq1$ is technically
  correct.}

@@

With the following code snippet, we perform weighted sampling from the
ROMC approximate posterior. Afterwards, we used some ELFI's built-in
tools to get a summary of the obtained samples. In figure
\ref{fig:example_sampling}, we observe the histogram of the weighted
samples and the acceptance region of the first deterministic function
(as before) alongside with the obtained samples obtained from
it. Finally, in the code snippet we demonstrate how to use the
\pinline{compute_expectation} function; in the current example we
define \pinline{h} in order to compute firstly the empirical mean and
afterwards the empirical variance. In both cases, the empirical result
is close to the ground truth $\mu = 0$ and $\sigma^2 = 1$.

\begin{pythoncode}
  seed = 21
  n2 = 50
  romc.sample(n2=n2, seed=seed)

@@

  print("Expected variance: %.3f" % romc.compute_expectation(h=lambda x: np.squeeze(x)**2))
  # Expected variance: 1.061
\end{pythoncode}

@@

\subsubsection*{Example - Evaluate Posterior}

The \pinline{romc.eval_unnorm_posterior(theta)} evaluates the
posterior at point $\theta$ using expression
\eqref{eq:aprox_posterior}. The \pinline{romc.eval_posterior(theta)}
approximates the partition function
$Z = \int_{\thetab} p_{d,\epsilon}(\thetab|\data) d\thetab$ using the
Riemann approximation as explained above. In our simple example, this
utility can provide a nice plot of the approximate posterior as
illustrated in figure~\ref{fig:approx_posterior}. We observe that the
approximation is quite close to the ground-truth posterior.

\begin{figure}[ht]
    \begin{center}
      \includegraphics[width=0.75\textwidth]{./latex_files/images/chapter3/example_posterior.png}
    \end{center}
  \caption[Approximate posterior evaluation, at the 1D example.]{Approximate posterior evaluation.}
  \label{fig:approx_posterior}
\end{figure}

>>>>

The inference part contains in four methods.

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.sample(n2)
----------------------------------------------------------------------------
\end{Code}

\noindent
This is the basic inference utility of the ROMC implementation; we
draw $n_2$ samples for each bounding box region. This gives a total of
$k \times n_2$, where $k < n_1$ is the number of the optimal points
remained after filtering\footnote{From the $n_1$ optimisation
  problems, only the ones with $g_i(\thetab_*) < \epsilon$ are
  maintained for building a bounding box}. The samples are drawn from
a uniform distribution $q_i$ defined over the corresponding bounding
box and the weight $w_i$ is computed as in
equation~\eqref{eq:sampling}.

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.compute_expectation(h)
----------------------------------------------------------------------------
\end{Code}

\noindent
This function computes the expectation
$E_{p(\thetab|\data)}[h(\thetab)]$ using
expression~\eqref{eq:expectation}. The argument \code{h} can be
any python \code{Callable}.

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.eval_unorm_posterior(theta, eps_cutoff=False)
----------------------------------------------------------------------------
\end{Code}

@@

\begin{Code}
---------------------------------- python ----------------------------------
>>> romc.eval_posterior(theta, eps_cutoff=False)
----------------------------------------------------------------------------
\end{Code}

\noindent
This function evaluates the normalised posterior estimating the
partition function $Z = \int p_{d,\epsilon}(\thetab|\data)d\thetab$;
with Riemann's integral approximation. The approximation is
computationally feasible only in low-dimensional parametric
spaces.

@@

In the following code snippet, we perform weighted sampling from the
ROMC approximate posterior. Afterwards, we used some of ELFI's
built-in tools to get a summary of the obtained samples. In figure
\ref{fig:example_sampling}, we observe the histogram of the weighted
samples and the acceptance region of the first deterministic function
(as before) alongside the samples obtained from it. Finally, in the
code snippet, we demonstrate how to use the \code{compute_expectation}
function; in the current example, we define \code{h} to compute
firstly the empirical mean and afterwards the empirical variance. In
both cases, the empirical result is close to the ground truth
$\mu = 0$ and $\sigma^2 = 1$. The
\code{romc.eval_unnorm_posterior(theta)} evaluates the posterior at
point $\theta$ using expression \eqref{eq:aprox_posterior}. The
\code{romc.eval_posterior(theta)} approximates the partition function
$Z = \int_{\thetab} p_{d,\epsilon}(\thetab|\data) d\thetab$ using the
Riemann approximation as explained above. In our simple example, this
utility can provide a nice plot of the approximate posterior as
illustrated in figure~\ref{fig:approx_posterior}. We observe that the
approximation is quite close to the ground-truth posterior.

\begin{Code}
------------------------------ python snippet ------------------------------  
  seed = 21
  n2 = 50
  romc.sample(n2=n2, seed=seed)

@@

  print("Expected variance: %.3f" % romc.compute_expectation(h=lambda x: np.squeeze(x)**2))
  # Expected variance: 1.061

  # eval unnorm posterior
  romc.eval_unnorm_posterior(theta=0)

  # check eval posterior
  romc.eval_posterior(theta=0)
----------------------------------------------------------------------------
\end{Code}

@@

\begin{figure}[h]
    \begin{center}
      \includegraphics[width=0.75\textwidth]{./latex_files/images/chapter3/example_posterior.png}
    \end{center}
  \caption[Approximate posterior evaluation, at the 1D example.]{Approximate posterior evaluation.}
  \label{fig:approx_posterior}
\end{figure}

### 2021-07-17 latex_files/ch3/3_2_4_evaluation.tex
added content in chapter 2
updated most content in chapter 2,3

The ROMC implementation provides two functions for evaluating the inference results,

\begin{enumerate}[label=(\roman*)]
\item \mintinline{python}{romc.compute_divergence(gt_posterior, bounds=None, step=0.1,}

      \mintinline{python}{                        distance="Jensen-Shannon")}

\item \mintinline{python}{romc.compute_ess()}
\end{enumerate}

\subsubsection*{Function (i): Compute the divergence between ROMC approximation and a ground-truth posterior}

\pinline{romc.compute_divergence(gt_posterior, bounds=None, step=0.1,}

\pinline{                     distance="Jensen-Shannon")}
\vspace{5mm}

\noindent
This function computes the divergence between the ROMC approximation
and the ground truth posterior. Since the computation is performed
using the Riemann's approximation, this method can only work in low
dimensional parameter spaces; it is suggested to be used for up to the
three-dimensional parameter space. As mentioned in the beginning of
this chapter, in a real-case scenario it is not expected the
ground-truth posterior to be available. However, in cases where the
posterior can be approximated decently well with a numerical approach
(as in the current example) or with some other inference approach,
this function can provide a numerical measure of the agreement
between the two approximations. The argument \pinline{step} defines
the step used in the Riemann's approximation and the argument
\pinline{distance} can take either the \pinline{Jensen-Shannon} or the
\pinline{KL-divergence} value, for computing the appropriate distance.

\subsubsection*{Function (ii): Compute the effective sample size of the weighted samples}

\mintinline{python}{romc.compute_ess()}
\vspace{5mm}

\noindent
This function computes the Effective Sample Size (ESS) using the
following expression~\autocite{Sudman1967},

@@

The ESS is a useful measure of the \textbf{actual} sample size, when
the samples are weighted. For example if in a population of $100$
samples one has a very large weight (e.g.\ $\approx 100$) whereas the
rest have small (i.e.\ $\approx 1$), the real sample size is close to
1; one sample dominates over the rest. Hence, the ESS provides a
measure of the equivalent uniformly weighted sample population.

\newpage

\begin{pythoncode}
res = romc.compute_divergence(wrapper, distance="Jensen-Shannon")                                 
print("Jensen-Shannon divergence: %.3f" % res)
# Jensen-Shannon divergence: 0.025

print("Nof Samples: %d, ESS: %.3f" % (len(romc.result.weights), romc.compute_ess()))
# Nof Samples: 19950, ESS: 16694.816
\end{pythoncode}

>>>>

The ROMC implementation provides two functions for evaluating the inference results.

\begin{Code}
---------------------------------- python ----------------------------------  
>>> romc.compute_divergence(gt_posterior,
                            bounds=None,
                            step=0.1,
                            distance="Jensen-Shannon")
----------------------------------------------------------------------------
\end{Code}

\noindent
This function computes the divergence between the ROMC approximation
and the ground truth posterior. Since the computation is performed
using the Riemann's approximation, this method can only work in low
dimensional parameter spaces; it is suggested to be used for up to the
three-dimensional parameter space. As mentioned at the beginning of
this chapter, in a real-case scenario it is not expected the
ground-truth posterior to be available. However, in cases where the
posterior can be approximated decently well with a numerical approach
(as in the current example) or with some other inference approach,
this function can provide a numerical measure of the agreement between
the two approximations. The argument \code{step} defines the step used
in the Riemann's approximation and the argument \code{distance} can
take either the \code{Jensen-Shannon} or the \code{KL-divergence}
value, for computing the appropriate distance.

\begin{Code}
---------------------------------- python ----------------------------------  
>>> romc.compute_ess()
----------------------------------------------------------------------------
\end{Code}

\noindent
This function computes the Effective Sample Size (ESS) using the
following expression~\cite{Sudman1967},

@@

The ESS is a valuable measure of the \textbf{actual} sample size when
the samples are weighted. For example, if in a population of $100$
samples one has an enormous weight (e.g.\ $\approx 100$) whereas the
rest have small (i.e.\ $\approx 1$), the actual sample size is close
to 1; one sample dominates over the rest. Hence, the ESS provides a
measure of the equivalent uniformly weighted sample population.

\begin{Code}
------------------------------ python snippet ------------------------------  
  res = romc.compute_divergence(wrapper, distance="Jensen-Shannon")                                 
  print("Jensen-Shannon divergence: %.3f" % res)
  # Jensen-Shannon divergence: 0.025

  nof_samples = len(romc.result.weights)
  ess = romc.compute_ess()
  print("Nof Samples: %d, ESS: %.3f" % (nof_samples, ess))
  # Nof Samples: 19950, ESS: 16694.816
----------------------------------------------------------------------------  
\end{Code}

### 2021-07-17 paper.tex
updated most content in chapter 2,3

%% Vasilis packages
\usepackage{amsmath,amssymb,bbold, xfrac}
\usepackage{xcolor}

% must be removed
\usepackage{multirow}
\usepackage{caption}

%% just to show content -> they should be replaced by the journals packages in the end
\usepackage{algorithm} 
\usepackage[noend]{algpseudocode}

%% just to show content -> they should be replaced by the journals packages in the end
% Python style code
\usepackage[finalizecache, cachedir=./minted_files]{minted}
\usemintedstyle{tango}
\usepackage{listings}
\usepackage{xcolor}
\newcommand{\pinline}[1]{\mintinline{python}{#1}}
\newminted{python}{framesep=2mm,
baselinestretch=1.2,
fontsize=\small,
frame=single
}

@@

% \begin{leftbar}
% The introduction is in principle ``as usual''. However, it should usually embed
% both the implemented \emph{methods} and the \emph{software} into the respective
% relevant literature. For the latter both competing and complementary software
% should be discussed (within the same software environment and beyond), bringing
% out relative (dis)advantages. All software mentioned should be properly
% \verb|\cite{}|d. (See also Appendix~\ref{app:bibtex} for more details on
% \textsc{Bib}{\TeX}.)

@@

%% - Virtually all JSS manuscripts list source code along with the generated
%%   output. The style files provide dedicated environments for this.
%% - In R, the environments {Sinput} and {Soutput} - as produced by Sweave() or
%%   or knitr using the render_sweave() hook - are used (without the need to
%%   load Sweave.sty).
%% - Equivalently, {CodeInput} and {CodeOutput} can be used.
%% - The code input should use "the usual" command prompt in the respective
%%   software system.
%% - For R code, the prompt "R> " should be used with "+  " as the
%%   continuation prompt.
%% - Comments within the code chunks should be avoided - these should be made
%%   within the regular LaTeX text.

% \begin{leftbar}
% For code input and output, the style files provide dedicated environments.
% Either the ``agnostic'' \verb|{CodeInput}| and \verb|{CodeOutput}| can be used
% or, equivalently, the environments \verb|{Sinput}| and \verb|{Soutput}| as
% produced by \fct{Sweave} or \pkg{knitr} when using the \code{render_sweave()}
% hook. Please make sure that all code is properly spaced, e.g., using
% \code{y = a + b * x} and \emph{not} \code{y=a+b*x}. Moreover, code input should
% use ``the usual'' command prompt in the respective software system. For
% \proglang{R} code, the prompt \code{"R> "} should be used with \code{"+  "} as
% the continuation prompt. Generally, comments within the code chunks should be
% avoided -- and made in the regular {\LaTeX} text instead. Finally, empty lines
% before and after code input/output should be avoided (see above).
% \end{leftbar}

@@

\section{ROMC module in ELFI}  % better title needed

\subsection{Engine for likelihood-free inference (ELFI)}
\input{latex_files/ch3/3_1_ELFI.tex}

\subsection{Design Principles}
\input{latex_files/ch3/3_2_design_principles.tex}

\subsection{Extensibility} % better title
\input{latex_files/ch3/3_3_extensibility.tex}

@@

%% -- Summary/conclusions/discussion -------------------------------------------

\section{Summary and discussion} \label{sec:summary}
lalala

@@

% The results in this paper were obtained using
% \proglang{R}~3.4.1 with the
% \pkg{MASS}~7.3.47 package. \proglang{R} itself
% and all packages used are available from the Comprehensive
% \proglang{R} Archive Network (CRAN) at
% \url{https://CRAN.R-project.org/}.

>>>>

%% Vasilis packages
\usepackage{amsmath,amssymb,bbold, xfrac}
\usepackage{xcolor}
\usepackage{multirow}
\usepackage{caption}
\usepackage{float}

%% maybe they should be replaced by the journals packages in the end
\usepackage{algorithm} 
\usepackage[noend]{algpseudocode}

@@

% \begin{leftbar}
% The introduction is in principle ``as usual''. However, it should usually embed
% both the implemented \emph{methods} and the \emph{software} into the respective
% relevant literature. For the latter both competing and complementary software
% should be discussed (within the same software environment and beyond), bringing
p% out relative (dis)advantages. All software mentioned should be properly
% \verb|\cite{}|d. (See also Appendix~\ref{app:bibtex} for more details on
% \textsc{Bib}{\TeX}.)

@@

% - Virtually all JSS manuscripts list source code along with the generated
%   output. The style files provide dedicated environments for this.
% - In R, the environments {Sinput} and {Soutput} - as produced by Sweave() or
%   or knitr using the render_sweave() hook - are used (without the need to
%   load Sweave.sty).
% - Equivalently, {CodeInput} and {CodeOutput} can be used.
% - The code input should use "the usual" command prompt in the respective
%   software system.
% - For R code, the prompt "R> " should be used with "+  " as the
%   continuation prompt.
% - Comments within the code chunks should be avoided - these should be made
%   within the regular LaTeX text.

\begin{leftbar}
For code input and output, the style files provide dedicated environments.
Either the ``agnostic'' \verb|{CodeInput}| and \verb|{CodeOutput}| can be used
or, equivalently, the environments \verb|{Sinput}| and \verb|{Soutput}| as
produced by \fct{Sweave} or \pkg{knitr} when using the \code{render_sweave()}
hook. Please make sure that all code is properly spaced, e.g., using
\code{y = a + b * x} and \emph{not} \code{y=a+b*x}. Moreover, code input should
use ``the usual'' command prompt in the respective software system. For
\proglang{R} code, the prompt \code{"R> "} should be used with \code{"+  "} as
the continuation prompt. Generally, comments within the code chunks should be
avoided -- and made in the regular {\LaTeX} text instead. Finally, empty lines
before and after code input/output should be avoided (see above).
\end{leftbar}

@@

\section{ROMC module in ELFI}  % better title needed
In this section, we present our implementation. At~\ref{subsec:ELFI}
we introduce the reader to the \proglang{Python} package \pkg{Engine
  for likelihood-free inference (ELFI)}
package. At~\ref{subsec:general_design} we explain the general design
that we follow. Then, we divide the implementation into three parts;
at~\ref{subsec:training} we present the training (fitting) part,
at~\ref{subsec:inference} the inference part and
at~\ref{subsec:evaluation} the evaluation part. Finally,
at~\ref{subsec:extensibility} we describe how a user may extend ROMC
with its custom modules.

\subsection{Engine for likelihood-free inference (ELFI)}
\label{subsec:ELFI}
\input{latex_files/ch3/3_1_ELFI.tex}

\subsection{Design principles} 
\label{subsec:general_design}
\input{latex_files/ch3/3_2_1_general_design}

\subsection{Training part} 
\label{subsec:training}
\input{latex_files/ch3/3_2_2_training}

\subsection{Inference part} 
\label{subsec:inference}
\input{latex_files/ch3/3_2_3_inference}

\subsection{Evaluation part} 
\label{subsec:evaluation}
\input{latex_files/ch3/3_2_4_evaluation}

\subsection{Extensibility} % better title
\label{subsec:extensibility}
\input{latex_files/ch3/3_3_extensibility.tex}

@@

%% -- Summary/conclusions/discussion -------------------------------------------
\section{Summary and discussion} \label{sec:summary}

@@

The results in this paper were obtained using
\proglang{R}~3.4.1 with the
\pkg{MASS}~7.3.47 package. \proglang{R} itself
and all packages used are available from the Comprehensive
\proglang{R} Archive Network (CRAN) at
\url{https://CRAN.R-project.org/}.
